Diffusive Homeostasis in a Self-Organizing Recurrent Neural Network \\ \begin{large} Spatially dependent Interaction as a Determinant of Neural Activity and Plasticity \end{large}

### 2017-03-08 draft.tex
.

\subsubsection{Synaptic Lifetimes}
Synaptic lifetimes have been shown to approximately follow a power law distribution in earlier versions of LIF-SORN and binary SORN \cite{SORN_Paper,Pengsheng_2013}. Power law distributions appear in different areas of natural and social sciences \cite{Gabaix_1999,Pascual_2002,Kagan_2010}. Some quantity $x$ is said to be power law distributed if its distribution $p(x)$ obeys
\begin{equation}
p(x) \propto x^{-\alpha}
\label{Pow_Law}
\end{equation}
with an exponent $\alpha>0$. Note that for this distribution to be properly normalized, one needs to set a lower (upper) bound $x_{\rm min.}$($x_{\rm max.}$) if $\alpha \geq 1$ ($\alpha \leq 1$). This family of distributions is of particular interest with respect to the fact that the ``long tail" of the distribution contains a significant portion of the total normalized integral of the probability density function. In other words, the probability of the appearance of exceptionally large values of $x$ can not be regarded as insignificant. Analyses of empirical data that is believed to obey a power-law distribution thus usually focus on the statistics of rare large events.

@@

Another question regarding the data

@@

 This feature was preserved under diffusive homeostasis, as shown in Figure \ref{Syn_Lifetimes}. The resulting slopes are within previously reported values of $\mathrm{\approx - 5/3}$ \cite{SORN_Paper}.
\begin{figure}
%%% plot_syn_lifetimes.py
\includegraphics[width=\textwidth]{../../plots/syn_topology/syn_lifetimes.png}
\caption{Distribution of lifetimes of recurrent excitatory synaptic connections. Linear fits ignoring falloff (only using bins with $\rm t \leq 150 \, s$) for large synaptic lifetimes resulted in a slope of $\mathrm{-1.87}$ for diffusive homeostasis and $\mathrm{-1.68}$ for non-diffusive homeostasis.}
\label{Syn_Lifetimes}
\end{figure}
To get an approximative explanation for the observed slope in the region of small lifetimes, we note that short synaptic lifetimes are most likely coming from synapses whose weights did not reach large values compared to the initial weight. The effect of multiplicative normalization is small compared to additive STDP in this regime of small weights. Small weights are therefore determined by additive quasi-random (due to their weak coupling to the postsynaptic neuron) fluctuations, which, as a very simple approximation, can be described as a Brownian motion with a certain standard deviation $\mathrm{\sigma}$. The first-hitting time of this Brownian motion is described by a Lévy distribution \cite{Roy_1968}:
\begin{equation}
p(t,w_0,w_{\rm prune}) = \frac{ | w_0-w_{\rm prune}|}{\sqrt{2\pi \sigma t^3}} \mathrm{exp} \left( \frac{ \left( w_0-w_{\rm prune} \right) ^2}{2\sigma t}\right)
\label{Levy_Dist}
\end{equation}
The term $\mathrm{t^{-3/2}}$ dominates for larger $\mathrm{t}$, which very roughly resembles the power-law exponents found in Figure \ref{Syn_Lifetimes}. Zheng et al. found a similar slope in the binary version of the SORN and also argued that this could be referred to random-walk behavior \cite{Pengsheng_2013}. Still, understanding the exponent of this distribution is subject to ongoing research and it has been argued that a more precise explanation could be used to relate certain parameters of synaptic plasticity to the value of this exponent. Especially, the balance between long-term potentiation and depression is thought to influence this exponent, where potentiation/depression dominated plasticity leads to a smaller/larger absolute value of the exponent of the distribution.

@@

A comparison of statistics generated with shuffled versions of the weight matrices illustrates that above-chance exceptionally strong mean outgoing weights were indeed present. To check for statistical significance, we performed a two-sample Kolmogorov-Smirnov tests between all datasets depicted in Figure \ref{Out_Weight_Mean_Quantile}. This statistical test allows to compare the empirical distribution functions of two sets of random variables by means of . More specifically, the null hypothesis states that both sets share the same underlying probability function. Thus, in our case, there should be a high statistical significance towards the rejection of this hypothesis. Though we did calculate the p-value for all possible pairs of datasets (28 in total), only a number of those were of relevance since we were interested in statistical differences due to diffusive homeostasis. These are listed in Table \ref{Kolmogorov_Mean_Out}. As the values indicate, the statistical significance is very high. Apart from the difference between diffusive and non-diffusive homeostasis, the difference between original and randomly shuffled versions of the weight matrices also appear to be significant.

>>>>

\subsubsection{Synaptic Lifetimes}\label{Section_Syn_Lifetimes}
Synaptic lifetimes have been shown to approximately follow a power law distribution in earlier versions of LIF-SORN and binary SORN \cite{SORN_Paper,Pengsheng_2013}. Power law distributions appear in different areas of natural and social sciences \cite{Gabaix_1999,Pascual_2002,Kagan_2010}. Some quantity $x$ is said to be power law distributed if its distribution $\mathrm{p}(x)$ obeys
\begin{equation}
\mathrm{p}(x) = C x^{-\alpha}
\label{Pow_Law}
\end{equation}
with an exponent $\alpha>0$ and $C$ an appropriate normalization. Note that for this distribution to be properly normalized, one needs to set a lower (upper) bound $x_{\rm min.}$($x_{\rm max.}$) if $\alpha \geq 1$ ($\alpha \leq 1$). This family of distributions is of particular interest with respect to the fact that the ``long tail" of the distribution contains a significant portion of the total normalized integral of the probability density function. In other words, the probability of the appearance of exceptionally large values of $x$ can not be regarded as insignificant. Analyses of empirical data that is believed to obey a power-law distribution thus usually focus on the statistics of rare large events.

@@

Another question we faced was which method of estimating the power-law exponent to use. Since taking the logarithm of Equation \eqref{Pow_Law} yields
\begin{equation}
\mathrm{ln}(\mathrm{p}) = \mathrm{ln}(C) - \alpha \mathrm{ln}(x) \; ,
\label{Pow_Law_Log} 
\end{equation}
the most naive approach one could choose is to generate a histogram of the lifetimes, transform it into a log-log space and estimate the exponent via the slope of a straight line acquired via linear regression. Though this might seem to be a straight-forward procedure, it has been argued that it can easily lead to wrong estimates due to large systematic errors, and common error estimates of the regression cannot be trivially interpreted \cite{Bauke_2007,Clauset_2009}. Clauset et al. have shown that the method of maximum likelihood estimation (MLE) provides a better framework for extracting power-law exponents \cite{Clauset_2009}. The basic idea of a MLE is to construct a family of probability density functions (PDF) $\rm f \left( \cdot| \left( \theta_0,\theta_1,... \right) \right)$ by means of one or more parameters $\theta_i$ one wishes to estimate \cite{EncyMath_Max_Like}. In the case of a power law, the family of PDFs would be characterized by the single exponent $\alpha$ and the function given in Equation \eqref{Pow_Law}. This parameter is then said to best characterize a given set of data points $(x_0,...,x_n)$ if its value $\alpha_{\rm max.}$ maximizes the joint probability of finding (or measuring) these data points under the assumption that they are independent and identically distributed according to $\rm f(x_i|\alpha)$. The expression one needs to maximize with respect to $\alpha$ is thus given by
\begin{equation}
\rm p_{\rm joint}(\alpha) = f(x_0|\alpha)f(x_1|\alpha)...f(x_n|\alpha) \; .
\label{Joint_Prob_Pow_Law}
\end{equation}
Ideally, an analytic expression for finding the maximum for the given set of data points exists. Otherwise, numerical optimization methods have to be used. For a continuous power-law PDF, such an expression indeed exists:\begin{equation}
\alpha_{\rm max.} = 1 + n \left[ \sum_{i=1}^n \mathrm{ln}\frac{x_i}{x_{\rm min.}} \right]^{-1}
\end{equation}
where $n$ is the number of data points and $x_{\rm min.}$ the smallest observed value \cite{Clauset_2009}. This equation does not apply for the discrete variant of a power-law distribution, where only integer values can be observed. Unfortunately, tracking synaptic connections in our network only allowed us to check the state of the connections once per second. Therefore, our dataset indeed only consists of integer multiples of one second. Numerical methods thus had to be used for proper estimation of the exponent.

We made use of the \textit{powerlaw python package} for this purpose \cite{Powerlaw_Package,Alstott_2014}. It provides a toolset for the analysis of potentially power-law distributed data, applying the maximum likelihood method as described by Clauset et al. \cite{Clauset_2009}. An additional feature of this package---also originally proposed by Clauset et al.---is the possibility to automatically find a lower bound $x_{\rm min.}$ below which data points are discarded from the analysis, based on the goodness of the estimated power law to the actual data. The authors argue that this feature accounts for the possibility that smaller values of a dataset might not as strictly follow a power law as found in the tail of the distribution. This optimization is based on the Kolmogorov-Smirnov statistic, which is the maximum absolute distance $d_{\rm max}$ between the empirical cumulative distribution function of the data and the cumulative distribution function of the analytic prediction. It can be further used to perform a statistical test, allowing a judgment whether a given dataset of random variables was generated from the given PDF \cite{EncyMath_Kolmogorov_Smirnov}. For the sake of finding the optimal value of $x_{\rm min.}$ however, the authors simply minimize $d_{\rm max}$ in order to find the best fitting setting.

@@

This feature was preserved under diffusive homeostasis, as shown in Figure \ref{Syn_Lifetimes}. The resulting slopes are within previously reported values of $\mathrm{\approx - 5/3}$ \cite{SORN_Paper}.
\begin{figure}
%%% plot_syn_lifetimes.py
\includegraphics[width=\textwidth]{../../plots/syn_topology/syn_lifetimes_new.png}
\caption{Distribution of lifetimes of recurrent excitatory synaptic connections. Linear fits ignoring falloff (only using bins with $\rm t \leq 150 \, s$) for large synaptic lifetimes resulted in a power-law exponent of $\mathrm{\alpha_{\rm Diff.} = -1.455}$ (standard error $\mathrm{\sigma_{\rm Diff.} = 5.52 \cdot 10^{-4}}$) for diffusive homeostasis and $\mathrm{\alpha_{\rm Non-Diff.} = -1.468}$ (standard error $\mathrm{\sigma_{\rm Non-Diff.} = 5.70 \cdot 10^{-4}}$) for non-diffusive homeostasis.}
\label{Syn_Lifetimes}
\end{figure}
To get an approximative explanation for the observed slope in the region of small lifetimes, we note that short synaptic lifetimes are most likely coming from synapses whose weights did not reach large values compared to the initial weight. The effect of multiplicative normalization is small compared to additive STDP in this regime of small weights. Small weights are therefore determined by additive quasi-random (due to their weak coupling to the postsynaptic neuron) fluctuations, which, as a very simple approximation, can be described as a Brownian motion with a certain standard deviation $\mathrm{\sigma}$. The first-hitting time of this Brownian motion is described by a Lévy distribution \cite{Roy_1968}:
\begin{equation}
p(t,w_0,w_{\rm prune}) = \frac{ | w_0-w_{\rm prune}|}{\sqrt{2\pi \sigma t^3}} \mathrm{exp} \left( \frac{ \left( w_0-w_{\rm prune} \right) ^2}{2\sigma t}\right)
\label{Levy_Dist}
\end{equation}
The term $\mathrm{t^{-3/2}}$ dominates for larger $\mathrm{t}$, which very roughly resembles the power-law exponents found in Figure \ref{Syn_Lifetimes}. Zheng et al. found a similar slope in the binary version of the SORN and also argued that this could be referred to random-walk behavior \cite{Pengsheng_2013}. Still, understanding the exponent of this distribution is subject to ongoing research and it has been argued that a more precise explanation could be used to relate certain parameters of synaptic plasticity to the value of this exponent. Especially, the balance between long-term potentiation and depression is thought to influence this exponent, where potentiation/depression dominated plasticity leads to a smaller/larger absolute value of the exponent of the distribution.

@@

A comparison of statistics generated with shuffled versions of the weight matrices illustrates that above-chance exceptionally strong mean outgoing weights were indeed present. To check for statistical significance, we performed a two-sample Kolmogorov-Smirnov test between all datasets depicted in Figure \ref{Out_Weight_Mean_Quantile}. It is based on the same principle as the one-sample Kolmogorov-Smirnov test used in Section \ref{Section_Syn_Lifetimes}. The only difference is that two empirical cumulative distributions are compared instead of comparing one empirical dataset to a given analytical function. The null hypothesis of the test states that both sets share the same underlying probability function. Thus, in our case, there should be a high statistical significance towards the rejection of this hypothesis. Though we did calculate the p-value for all possible pairs of datasets (28 in total), only a number of those were of relevance since we were interested in statistical differences due to diffusive homeostasis. These are listed in Table \ref{Kolmogorov_Mean_Out}. As the values indicate, the statistical significance is very high. Apart from the difference between diffusive and non-diffusive homeostasis, the difference between original and randomly shuffled versions of the weight matrices also appear to be significant.
